\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-04-09 paper.qmd
revision of kD model

bibliography: bibliography.bib  
header-includes: | 
  \usepackage{amsmath}
  \usepackage{float}
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \usepackage{bm}
  \def\tightlist{}
  \usepackage{setspace}
  \newcommand\pD{$p\text{-}D$\ }
  \newcommand\kD{$k\text{-}D$\ }
  \newcommand\dD{$d\text{-}D$\ }
---

@@

At first glance, thinking of NLDR as a model fitted to the data might seem strange. It is a model in the sense that it is a "a simplified representation or abstraction of a system, process, or phenomenon in the real world". The \pD observations are the realization of the phenomenon, and the \kD NLDR layout is the simplified representation. From a statistical perspective we can consider the distances between points in the \kD layout to be variance that the model explains, and the (relative) difference with their distances in \pD is the error, or unexplained variance. Abstractly, we can also imagine that the positioning of points in 2D represents fitted values, that will have some prescribed position in \pD that can be compared with the observed value. 
<!--### Notation -->

@tbl-notation summarises the notation used to explain the new methodology. The observed data is denoted as $\mathbfit{x}_{n \times p}$ where $x_{ij}$ would indicate the $i^{th}$ observation on the $j^{th}$ variable sampled from a population $\mathbfit{X}$. To refer to variable $j$, we would use $X_j$.

@@

We define the NLDR model as a function that maps \pD to \kD, and takes some (hyper)parameters $\vec{\theta}$. That is,

$$g : \mathbb{R}^p \rightarrow \mathbb{R}^k.$$

The parameter vector, $\vec{\theta}$, depends on the choice of $g$, and can be considered part of 'model fitting' in the traditional sense. Common choices for $g$ include tSNE, UMAP, PHATE, TriMAP, or PaCMAP, although in theory any function that maps the entire space $\mathbb{R}^p$ to a subset of $\mathbb{R}^k$ is suitable. Any input requirements for the data (such as normalization, or preprocessing through the use of PCA or similar) is considered part of the function $g$.

Next, we create a model of the \kD representation of the data using hexagonal binning. The aim of this process is to reduce the noise in the original data that may be preserved by $g$. Define a hexagonal grid across the \kD space, defined by the height and width of the hexagon ($h_b$ and $w_b$ respectively), the aspect ratio of the hexagon ($h_r$), and the starting coordinates of the hexagonal grid $(x_s, y_s)$. We deliberately separate out the creation of the hexagonal grid from the mapping of points on the grid.

With the grid defined, the \kD points will belong to a hexagonal bin. That is, for each $y \in \mathbfit{Y}$, we can (uniquely) identify the hexagon that the point belongs to. Let,

$$h : \mathbb{R}^k \rightarrow \mathbb{R}^k, $$

be the function that maps a point in \kD space to it's hexagonal centroid. It follows that $h \circ g : \mathbb{R}^p \rightarrow \mathbb{R}^k$ maps each point in $\mathbfit{X}$ to a \kD centroid.

@@

- Summarise contributions
- Explain where it is expected or not expected to work, eg higher dimensional relationships
- Human behaviour, the desire to have more certainty, and a tendency to prefer the well-separated views
- Diagnostic app to explore differences in distances
- What might be useful enhancements

>>>>

bibliography: bibliography.bib  
header-includes: | 
  \usepackage{amsmath}
  \usepackage{float}
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \usepackage{bm}
  \def\tightlist{}
  \usepackage{setspace}
  \newcommand\pD{$p\text{-}D$}
  \newcommand\kD{$k\text{-}D$}
  \newcommand\dD{$d\text{-}D$}
---

@@

At first glance, thinking of NLDR as a modeling technique might seem strange. It is a simplified representation or abstraction of a system, process, or phenomenon in the real world. The \pD   observations are the realization of the phenomenon, and the \kD   NLDR layout is the simplified representation. From a statistical perspective we can consider the distances between points in the \kD   layout to be variance that the model explains, and the (relative) difference with their distances in \pD   is the error, or unexplained variance. We can also imagine that the positioning of points in 2D represent the fitted values, that will have some prescribed position in \pD   that can be compared with their observed values. This is the conceptual framework underlying the more formal versions of factor analysis [@cfa69] and multidimensional scaling (MDS) [@borg2005]. (Note that, for this thinking the full \pD   data needs to be available, not just the interpoint distances.)
<!--### Notation -->

<!-- @tbl-notation summarises the notation used to explain the new methodology. The observed data is denoted as $\mathbfit{x}_{n \times p}$ where $x_{ij}$ would indicate the $i^{th}$ observation on the $j^{th}$ variable sampled from a population $\mathbfit{X}$. To refer to variable $j$, we would use $X_j$.-->

@@

We define the NLDR model as function $g\text{:}~ \mathbb{R}^p \rightarrow \mathbb{R}^k$, with (hyper-)parameters $\mathbfit{\theta}$. The parameters, $\mathbfit{\theta}$, depend on the choice of $g$, and can be considered part of model fitting in the traditional sense. Common choices for $g$ include functions used in tSNE, UMAP, PHATE, TriMAP, PaCMAP, or MDS, although in theory any function that maps the entire space $\mathbb{R}^p$ to a subset of $\mathbb{R}^k$ is suitable. <!--Any input requirements for the data (such as normalization, or preprocessing through the use of PCA or similar) is considered part of the function $g$.-->

With our goal being to make a representation of this 2D layout that can be lifted into high-dimensional space, the layout needs to be augmented to include neighbour information. A simple approach would be to triangulate the points and add edges. A more stable approach is to first hexagonally bin the data, and connect the bin centroids. This process serves reduce some noise in the resulting surface shown in \pD.

Define a hexagonal grid across the \kD space, using hexagons with fixed height ($h_b$), width ($w_b$). The starting coordinates $(x_s, y_s)$ mark the lower left of the grid. <!--We deliberately separate out the creation of the hexagonal grid from the mapping of points on the grid.-->
Each of the \kD points will belong to a hexagonal bin. That is, for each $y \in \mathbfit{Y}$, we can (uniquely) identify the hexagon that the point belongs to. Let $h : \mathbb{R}^k \rightarrow \mathbb{R}^k$ be the function that maps a point in \kD space to it's hexagonal centroid ($C^{(k)}$). <!--It follows that $h \circ g : \mathbb{R}^p \rightarrow \mathbb{R}^k$ maps each point in $\mathbfit{X}$ to a \kD centroid.-->

@@

- Summarise contributions
- Explain where it is expected or not expected to work, eg higher dimensional relationships
- Human behaviour, the desire to have more certainty, and a tendency to prefer the well-separated views
- Predicting new observations in $k$-D
- Extending layouts beyond $k$-D, when 2D is clearly inadequate.
- Diagnostic app to explore differences in distances
- What might be useful enhancements

### 2024-04-09 paper.tex
Change normalisation spelling to American
revision of kD model

\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-1in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{1.7in}%
\addtolength{\topmargin}{-1in}%
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{amsmath}
\usepackage{float}
\usepackage{hyperref}
\usepackage[utf8]{inputenc}
\usepackage{bm}
\def\tightlist{}
\usepackage{setspace}
\newcommand\pD{$p\text{-}D$\ }
\newcommand\kD{$k\text{-}D$\ }
\newcommand\dD{$d\text{-}D$\ }
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{agsm}
\usepackage{bookmark}

@@

At first glance, thinking of NLDR as a model fitted to the data might
seem strange. It is a model in the sense that it is a ``a simplified
representation or abstraction of a system, process, or phenomenon in the
real world''. The \pD observations are the realization of the
phenomenon, and the \kD NLDR layout is the simplified representation.
From a statistical perspective we can consider the distances between
points in the \kD layout to be variance that the model explains, and the
(relative) difference with their distances in \pD is the error, or
unexplained variance. Abstractly, we can also imagine that the
positioning of points in 2D represents fitted values, that will have
some prescribed position in \pD that can be compared with the observed
value.

Table~\ref{tbl-notation} summarises the notation used to explain the new
methodology. The observed data is denoted as
\(\mathbfit{x}_{n \times p}\) where \(x_{ij}\) would indicate the
\(i^{th}\) observation on the \(j^{th}\) variable sampled from a
population \(\mathbfit{X}\). To refer to variable \(j\), we would use
\(X_j\).

@@

\begin{tabular}{>{\raggedright\arraybackslash}p{3cm}>{\raggedright\arraybackslash}p{12cm}}
\toprule
\textbf{Notation} & \textbf{Description}\\
\midrule
$n, p, k$ & number of observations, variables, embedding dimension, respectively\\
$\mathbfit{X}, \mathbfit{x}$ & $p$-dimensional data (population, sample)\\
$\mathbfit{y}$ & $k$-dimensional layout\\
$P$ & orthonormal basis, generating a $d\text{-}dimensional$ linear projection of $p$-dimensional data\\
$T$ & true  model\\
\addlinespace
$y_\text{max}$ & maximum value of the scaled second embedding component\\
$r_1$, $r_2$ & range of the first and  second embedding component\\
$h_b$, $w_b$ & height and width of the hexagon\\
$a_r$ & aspect ratio of the 2D layout before scaling\\
$h_r$ & hexagon ratio\\
\addlinespace
$s$ & hexagonal size (radius of the outer circle surrounding the hexagon)\\
$(x_s, \ y_s)$ & starting coordinates of the hexagonal grid\\
$q_x$, $q_y$ & buffer amount along the x and y axes\\
$h_s$, $v_s$ & horizontal and vertical distance between adjacent hexagon bin centroids\\
$d_x$, $d_y$ & horizontal spacing and adjusted vertical spacing for hexagonal coordinates\\
\addlinespace
$b_x$, $b_y$ & number of hexagon bins along the x and y axes\\
$b$, $b'$ & total and non-empty hexagon bins in the grid\\
$n_k$ & number of observations within the $k^{th}$ hexagon\\
$(h^k_{x_i}, h^k_{y_i})$ & hexagonal grid coordinates of the $k^{th}$ hexagon\\
$C^{(j)}$ & $j$-dimensional bin centers\\
\bottomrule
\end{tabular}
\endgroup{}

@@

@@

We need to have some math that gives precision to the paragraphs in the
previous section, defining mapping from \pD to \kD to get the
\kD layout. Might be written below, too (XXX Michael??)

\subsection{\texorpdfstring{Generating the
\kD layout}{Generating the layout}}\label{generating-the-layout}

The usual procedures are followed to produce a \kD layout, using the
preferred method, t-SNE, UMAP, \ldots{} The data may need to be
pre-processed by standardising or transforming to principal components.

\subsection{Constructing the 2D model}\label{constructing-the-2d-model}

@@

@@

\begin{itemize}
\tightlist
\item
  Summarise contributions
\item
  Explain where it is expected or not expected to work, eg higher
  dimensional relationships
\item
  Human behaviour, the desire to have more certainty, and a tendency to
  prefer the well-separated views
\item
  Diagnostic app to explore differences in distances
\item
  What might be useful enhancements
\end{itemize}

>>>>

\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-1in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{1.7in}%
\addtolength{\topmargin}{-1in}%
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{amsmath}
\usepackage{float}
\usepackage{hyperref}
\usepackage[utf8]{inputenc}
\usepackage{bm}
\def\tightlist{}
\usepackage{setspace}
\newcommand\pD{$p\text{-}D$}
\newcommand\kD{$k\text{-}D$}
\newcommand\dD{$d\text{-}D$}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{agsm}
\usepackage{bookmark}

@@

At first glance, thinking of NLDR as a modeling technique might seem
strange. It is a simplified representation or abstraction of a system,
process, or phenomenon in the real world. The \pD   observations are the
realization of the phenomenon, and the \kD   NLDR layout is the
simplified representation. From a statistical perspective we can
consider the distances between points in the \kD   layout to be variance
that the model explains, and the (relative) difference with their
distances in \pD   is the error, or unexplained variance. We can also
imagine that the positioning of points in 2D represent the fitted
values, that will have some prescribed position in \pD   that can be
compared with their observed values. This is the conceptual framework
underlying the more formal versions of factor analysis \citep{cfa69} and
multidimensional scaling (MDS) \citep{borg2005}. (Note that, for this
thinking the full \pD   data needs to be available, not just the
interpoint distances.)

@@

\begin{tabular}{>{\raggedright\arraybackslash}p{3cm}>{\raggedright\arraybackslash}p{12cm}}
\toprule
\textbf{Notation} & \textbf{Description}\\
\midrule
$n, p, k$ & number of observations, variables, embedding dimension, respectively\\
$\mathbfit{X}, \mathbfit{x}$ & $p$-dimensional data (population, sample)\\
$\mathbfit{y}$ & $k$-dimensional layout\\
$P$ & orthonormal basis, generating a $d\text{-}dimensional$ linear projection of $p$-dimensional data\\
$T$ & true  model\\
\addlinespace
$g$ & functional mapping from \pD to \kD, especially as prescribed by NLDR\\
$\mathbfit{\theta}$ & (Hyper-) parameters for NLDR method\\
$y_\text{max}$ & maximum value of the scaled second embedding component\\
$r_1$, $r_2$ & range of the first and  second embedding component\\
$h_b$, $w_b$ & height and width of the hexagon\\
\addlinespace
$a_r$ & aspect ratio of the 2D layout before scaling\\
$h_r$ & hexagon ratio\\
$s$ & hexagonal size (radius of the outer circle surrounding the hexagon)\\
$(x_s, \ y_s)$ & starting coordinates of the hexagonal grid\\
$q_x$, $q_y$ & buffer amount along the x and y axes\\
\addlinespace
$h_s$, $v_s$ & horizontal and vertical distance between adjacent hexagon bin centroids\\
$d_x$, $d_y$ & horizontal spacing and adjusted vertical spacing for hexagonal coordinates\\
$b_x$, $b_y$ & number of hexagon bins along the x and y axes\\
$b$, $b'$ & total and non-empty hexagon bins in the grid\\
$n_k$ & number of observations within the $k^{th}$ hexagon\\
\addlinespace
$(h^k_{x_i}, h^k_{y_i})$ & hexagonal grid coordinates of the $k^{th}$ hexagon\\
$C^{(j)}$ & $j$-dimensional bin centers\\
\bottomrule
\end{tabular}
\endgroup{}

@@

We define the NLDR model as function
\(g\text{:}~ \mathbb{R}^p \rightarrow \mathbb{R}^k\), with
(hyper-)parameters \(\mathbfit{\theta}\). The parameters,
\(\mathbfit{\theta}\), depend on the choice of \(g\), and can be
considered part of model fitting in the traditional sense. Common
choices for \(g\) include functions used in tSNE, UMAP, PHATE, TriMAP,
PaCMAP, or MDS, although in theory any function that maps the entire
space \(\mathbb{R}^p\) to a subset of \(\mathbb{R}^k\) is suitable.

With our goal being to make a representation of this 2D layout that can
be lifted into high-dimensional space, the layout needs to be augmented
to include neighbour information. A simple approach would be to
triangulate the points and add edges. A more stable approach is to first
hexagonally bin the data, and connect the bin centroids. This process
serves reduce some noise in the resulting surface shown in \pD.

Define a hexagonal grid across the \kD space, using hexagons with fixed
height (\(h_b\)), width (\(w_b\)). The starting coordinates
\((x_s, y_s)\) mark the lower left of the grid. Each of the \kD points
will belong to a hexagonal bin. That is, for each
\(y \in \mathbfit{Y}\), we can (uniquely) identify the hexagon that the
point belongs to. Let \(h : \mathbb{R}^k \rightarrow \mathbb{R}^k\) be
the function that maps a point in \kD space to it's hexagonal centroid
(\(C^{(k)}\)).

@@

@@

The final step of the algorithm is \emph{lifting} the \kD model back to
\pD, completing the \emph{model-in-the-data-space} picture. For this,
define the set \(H^k\) to be the set of points that belong to centroid
\(k\). That is, \(H^k = \{ y \; | \; g(y) = (h_x^k, h_y^k)\}\). We use
the \pD Euclidean mean of the points in \(H^k\) to map the centroid
\((h_x^k, h_y^k)\) to a point in \pD. Let the \(i\)-th component of the
\pD mean be

\[f_i = \frac{1}{\left|H^k\right|}\sum_{y \in H^k} y_i,\]

with corresponding \pD vector \(\vec{f_i}\). Then, \(f(h\circ g)(x)\)
maps a \pD point to the \pD model estimate, completing the model cycle.

@@

\begin{itemize}
\tightlist
\item
  Summarise contributions
\item
  Explain where it is expected or not expected to work, eg higher
  dimensional relationships
\item
  Human behaviour, the desire to have more certainty, and a tendency to
  prefer the well-separated views
\item
  Predicting new observations in \(k\)-D
\item
  Extending layouts beyond \(k\)-D, when 2D is clearly inadequate.
\item
  Diagnostic app to explore differences in distances
\item
  What might be useful enhancements
\end{itemize}
